Predictive modeling using genomic annotations

### 2018-10-01 Manuscript/manuscript.Rmd
updates
manuscript updates

###Methods 
We propose a novel data pre-processing pipeline that addresses those shortcomings. It includes an elastic-net regularized classification model, where the mixing proportion, $\alpha$, and the penalization parameter, $\lambda$, are tuned for over a grid of values. A number of classifier performance metrics was assessed, including the F1 measure and Matthew Correlation Coefficient (MCC).

###Results 
Data preprocessing (log2-transformation and standardization) improves the performance of the models. The elastic-net model outperformed multiple logistic regression models, with and without LASSO-regularized coefficients. Likewise, the variable importance of the elastic-net yielded more stable and biologically meaningful results compared to the variables selected by the multiple logistic regression models.

###Conclusions 
Current methods used to model the epigenomic features associated with TAD boundaries are not robust to handle properties of genomic data. Models applied to unprocessed data can have poor predictive performances. Typical performance assessment metrics, such as AUROC, can mask poor performance of the models; thus, the use of more balanced metrics, such as F1 and MCC, is warranted. The elastic-net model, with its dual use of the $l_1$ and $l_2$ regularization terms allows for both the removal of uninformative predictors, while being able to handle multiple correlated epigenomic features. Our model results in better performances and more accurate identification of important features associated with the formation of TAD boundaries.

@@

Due to the size of Hi-C data and the abundance of available genomic features, few methods have been developed to study the role and interplay of different sets of these features on the folding of DNA. Furthermore, many accepted methods ignore key characteristics of the data that, when addressed, may improve predictive performances. One such pitfall is the issue of highly imbalanced classes. The commonality between most accepted methods for predicting TAD boundaries considers binning the genome and labeling the bins as either containing a TAD boundary or not. However, due to the sparcity of TADs throughout the genome, most bins will not contain a TAD boundary. This will contribute to highly imbalanced classes and will also introduce predictors with near-zero variance. It has been shown that imbalanced classes can result in poor predictive performances of classification algorithms and unless the class imbalance is mild, researchers should always use an appropriate method for dealing with this (Blagus et al).  Furthermore, the inclusion of predictors with near-zero variance can add unnecessary noise to the model and decrease computational efficiency of predictive algorithms. Another problem lies in identifying which positional characteristics of the genomic features will optimize prediction. It is unclear if distances or overlaps between genomic bins and regions of functional elements contain the most information and will lead to better predictive performances. Lastly, given that some features tend to colocalize near TAD boundaries, most predictive models may be unable to identify the complete set of features most associated with boundary formation. This can lead to less informative results and restrict our knowledge of how these genomic elements correlate with eachother.

In a previous study (Mourad et al) both a multiple logistic regression model and a LASSO regularized regression model were applied to a wide array of architectural proteins from the GM12878 cell line to predict TAD boundaries. The set of features consisted of whether or not the coordinates of a functional element overlaped with a particular genomic bin (yes or no). In the case of a partial overlap, a percentage of overlap was used. However, the class imbalance issue was not addressed and neither model used is suited to handle multiple correlated predictors. Additionally, ROC curves and AUCs were used to evaluate model performance, which is problematic, especially for highly imbalaned classes (Hanczar et al). A better indicator of predictive performance with classification algorithms is to consider the F1 score and Matthew's Correlation Coefficient (MCC). In another study (Huang et al), a more robust model in the form of a Bayesian Additive Regression Trees (BART) model was used. The set of features consisted of a score that averaged the number of sequence reads for each functional element over a particular bin for the GM12878 cell line. Here, the class proportion was manually created by dividing the positive and negative classes into two equal subsets used for model training and testing. However, it is unclear what balancing technique was used. In addition to this, the model was only performed on a small set of histone marks while ignoring wider classes of epigenomic elements. Lastly, in a more recent study (Hong et al), a bayesian rigde regression model was iteratively used on a set of transcription factor binding sites, transcription start sites, and DNase I hypersensitive sites from the hESC cell line until the best combination of features led to the optimal prediction of TAD boundaries.  Here, the features of the model consisted of the number of occurences of each genomic element within each genomic bin. Balanced classes were manually created by randomly sampling from the set of bins that did not contain a TAD boundary. Given the structure of the search algorithm proposed, though, it can be computationally inefficient to have to iterate through a large set of genomic features to find the particular set that contributes to the best prediction. Because of that, only a maximum set of 10 genomic features were considered at any given time.

The types of models and the feature characteristics used in the aforementioned methods vary considerably. Because of the different cell lines available for analysis and the widening list of functional genomic elements, it can be quite useful to have a unified approach in predicting the formation of TAD boundaries. Therefore, in this study we have developed a novel data pre-processing pipeline that addresses the irregularities in the TAD boundary data provided by contact matrices at different resolutions for two cell lines, GM12878 and K562. We then performed a variety of machine learning algorithms on the pre-processes data to identify the genomic elements that were most predictive of TAD boundaries. In our models we incorporated the distances from bin to region as features. We found that the models performed on our pre-processed data performed better than previously established methods. Additionally, we were able to confirm that the cis-regulatory elements like insulators and promotors were highly predictive of TAD boundaries. The specific transcriptional repressor, CCCTC-binding factor (CTCF) was found to be highly predictive of TAD boundaries. Other non-transcription factor sites like DNase I hypersensitive sites and certain histone modifications like H3K36me3 were also found to be better predictors.  Our findings confirm that models with pre-processed data and distance-related genomic features not only perform better, but are also able to correctly identify a range of different functional genomic elements associated with the formation of TAD boundaries. This knowlege will allow us to gain better insight into the role that the spatial organisation of the human genome has on gene regulation.

@@

Publicly available topologically associating domain data was obtained from GEO with accession GSE53525 for the GM12878 and K562 cell lines.  The domain data was constructed from in situ Hi-C contact matrices at a 5kb and 10 kb resolution for the GM12878 and k562 cell lines respectively. The Arrowhead algorithm was used to identify TAD boundaries for a given contact matrix (Rao et. al). Identified TAD boundaries were represented by their genomic coordinates (hg19 human genome assembly), including chromosome, start and end coordinates. The start and end coordinates were concatenated, sorted, and unique coordinates representing unique TAD boundaries were obtained [??? Check for accuracy; !!! what do you mean by accuracy here?]. A depiction of the data is presented in Figure X.

The establishment of the data for modeling first consisted of binning the genome into equal sized intervals. This was performed at the same resolution as the respective cell line. That is, for GM12878, 5 kb genomic bins were used, while 10 kb bins were used for K562. Next, for the domain data provided by GM12878, the boundary coordinates were flanked on either side by 5 kb. For the K562, the boundary coordinates were flanked by 10 kb. In either case, a genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing an boundary (Y=0) (Figure X).

@@

Annotation data was obtained from the Encyclopedia of DNA Elements (ENCODE) Consortium.  These annotations consisted of genomic coordinates and were used to build the feature space, X = {$X_{1}$, . . ., $X_{p}$}, of the subsequent models. A full list of the genomic elements is provided in Table X. For each genomic element, the distance from the center of a genomic bin to the center of the nearest region (in base pairs) was calculated and included in the model.  If the center of the region given by a particular functional element was located within a genomic bin, then the distance was given to be 0. Additionally, a variable denoting the number of overlaps between a genomic bin and region representing a genomic element was totaled and included in the model (Figure X).

@@

We developed a pipeline that systematically evaluates the problems associated with TAD boundary data in a predictive modeling setting (Figure X). The pipeline was used to address issues related to normalization of predictors, class imbalance, and variable reduction. Prior to initiating the pipeline, predictors with near-zero variance were removed. To identify said predictors, the nearZeroVar function in the CARET package was used. The function considers two metrics. First, the ratio of the most frequent value over the second most frequent value was calculated. Ideally, this ratio would be close to one for well-behaved predictors. Next, the number of unique values divided by the total number of samples multiplied by 100 was calculated. Here, if the frequency ratio was greater than 20 and the percentage of unique values was less than 10, the predictor was flagged as having near-zero variance and removed.

At each subsequent stage in the pipeline, Elastic-Net models were performed and performance metrics such as F1 scores and MCCs were evaluated. The Elastic-Net model was used here because of its ability to be tuned and its computational speed relative to other machine learning algorithms. To avoid overfitting and more accurately assess performance, 10-fold cross validation was used for all models. For each elastic-net model the mixing parameter, alpha, was tuned over a grid of 10 evenly spaced values from 0 and 1. The penalty term, labmda, was tuned over a grid of 100 evenly spaced values from 0.01 to 100.

### NORMALIZATION

Distances on a genomic scale are frequently highly skewed. In order to correct for this, a log base 2 transformation was considered. Additionally, the predictors were standardized to have unit variance in order to be able to directly compare magnitudes of parameter coefficients.

### CLASS IMBALANCE

When binning the genome, the sparcity of TAD boundaries through the genome contributed to heavily imbalanced classes. That is, proportionally, only a small number of genomic bins overlapped with a flanked TAD boundary. To assess this issue, two methods were used. The first consisted of incorporating the SMOTE command from the DmWR package in R. SMOTE stands for Synthetic Minority Over-sampling Technique. The SMOTE algorithm incorporates both under-sampling (percent under) of the majority class and over-sampling (percent over) of the minority class. The minority class is over-sampled by taking each minority class sample and introducing synthetic examples along the line segments joining any/all of the k minority class nearest neighbors (here, k=5). Then, the majority class is under sampled by taking a random sample that matches some pre-specified percentage of the updated minority class. It is important to note that using a combination of 100/200 (percent over/percent under-sample) will always result in perfectly balanced classes. A total of four different combinations of percent over and under-sampling were used which included: 100/200, 200/200, 300/200, and 400/200. As mentioned before, the 100/200 combination created balanced classes, while the others create an unbalanced dataset with the negative class as the majority to reflect the proportionality of the classes in the original data. The second method involved iteratively under-sampling 100 time from the majority class to match the number in the minority class. For each sample, an elastic-net model was tuned and evaluated. Then row-wise means of the performance metrics for each of the 100 interations were taken to evaluate the models (Figure X).

>>>>

###Methods 
We propose a novel data pre-processing pipeline that addresses those shortcomings. A number of classifier performance metrics were assessed, including the F1 measure and Matthew Correlation Coefficient (MCC), and area under the ROC curve (AUROC).

###Results 
Data preprocessing (log2-transformation and standardization) improves the performance of classification algorithms and allows for the ability to more accurately predict which genomic features are most associated with TAD boundaries.

###Conclusions 
Current methods used to model the epigenomic features associated with TAD boundaries are not robust to handle properties of genomic data. Models applied to unprocessed data can have poor predictive performances. Focusing solely on typical performance assessment metrics, such as AUROC, can mask poor performance of the models; thus, the use of more balanced metrics, such as F1 and MCC, is warranted. Our model results in better performances and more accurate identification of important features associated with the formation of TAD boundaries.

@@

Due to the size of Hi-C data and the abundance of available genomic features, few methods have been developed to study the role and interplay of different sets of these features on the folding of DNA. Furthermore, many accepted methods ignore key characteristics of the data that, when addressed, may improve predictive performances. One such pitfall is the issue of highly imbalanced classes. The commonality between most accepted methods for predicting TAD boundaries considers binning the genome and labeling the bins as either containing a TAD boundary or not. However, due to the sparcity of TADs throughout the genome, most bins will not contain a TAD boundary. This will contribute to highly imbalanced classes and will also introduce predictors with near-zero variance. It has been shown that imbalanced classes can result in poor predictive performances of classification algorithms and unless the class imbalance is mild, researchers should always use an appropriate method for dealing with this (Blagus et al).  Furthermore, the inclusion of predictors with near-zero variance can add unnecessary noise to the model and decrease computational efficiency of predictive algorithms. Another problem lies in identifying which positional characteristics of the genomic features will optimize prediction. The use of distances (in base pairs) from genomic bins to regions of functional elements has largely been ignored. However, these distance characteristics offer a more accurate spatial representation of which genomic features are associated with TAD boundaries and have the potential to improve predictive performance. Lastly, given that some features tend to colocalize near TAD boundaries, most predictive models may be unable to identify the complete set of features most associated with boundary formation. This can lead to less informative results and restrict our knowledge of how these genomic elements correlate with eachother.

In a previous study (Mourad et al) both a multiple logistic regression model and a LASSO regularized regression model were applied to a wide array of architectural proteins from the GM12878 cell line to predict TAD boundaries. The set of features consisted of whether or not the coordinates of a functional element overlaped with a particular genomic bin (yes or no). In the case of a partial overlap, a percentage of overlap was used. However, the class imbalance issue was not addressed and neither model used is suited to handle multiple correlated predictors. Additionally, ROC curves and AUCs were used to evaluate model performance, which is problematic, especially for highly imbalaned classes (Hanczar et al). A better indicator of predictive performance with classification algorithms is to consider the F1 score and Matthew's Correlation Coefficient (MCC) as they involve all quadrants of a confusion matrix.

In this study we have developed a novel data pre-processing pipeline that addresses the irregularities in the TAD boundary data provided by contact matrices at different resolutions for two cell lines, GM12878 and K562. We then performed a variety of machine learning algorithms on the pre-processed data to identify the genomic elements that were most predictive of TAD boundaries. Firstly, we found that the models performed on balanced data out performed those on unbalanced data. Secondly, it was determined that multiple iterations of random under-sampling was the balancing technique that yielded the best results. Thirdly, the inclusion of distances as predictors was found to improve model performance. Lastly, it was found that performing an inverse normal transformation of the distance predictors further improved model performance.

After applying (insert final model(s)) we were able to confirm that cis-regulatory elements like insulators and promotors were highly predictive of TAD boundaries. The specific transcriptional repressor, CCCTC-binding factor (CTCF), was also found to be highly associated with TAD boundaries. Other non-transcription factor sites like DNase I hypersensitive sites and certain histone modifications like H3K36me3 were also found to be better predictors of TAD boundaries.  Our findings confirm that models with pre-processed data and distance-related genomic features not only perform better, but are also able to correctly identify a range of different functional genomic elements associated with the formation of TAD boundaries. This knowlege will allow us to gain better insight into the role that the spatial organisation of the human genome has on gene regulation.

@@

Publicly available topologically associating domain data was obtained from GEO with accession GSE53525 for the GM12878 and K562 cell lines (Rao et. al).  The domain data was constructed from in situ Hi-C contact matrices at a 5kb and 10 kb resolution for the GM12878 and k562 cell lines respectively. Identified TAD boundaries were represented by their genomic coordinates (hg19 human genome assembly), including chromosome, start and end coordinates. The start and end coordinates were concatenated, sorted, and unique coordinates representing the borders demarcating TADs were obtained.

The establishment of the data for modeling first consisted of binning the genome into equal sized intervals. This was performed at the same resolution as the respective cell line. That is, for GM12878, 5 kb genomic bins were used, while 10 kb bins were used for K562. Next, for the domain data provided by GM12878, the boundary coordinates were flanked on either side by 5 kb. For the K562, the boundary coordinates were flanked by 10 kb. In either case, a genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing an boundary (Y=0). An illustration of the data is presented in Figure X.

@@

Annotation data, in the form of functional genomic coordinates, from ChIP-seq experiments was obtained from the Encyclopedia of DNA Elements (ENCODE) Consortium. Similar to Mourad et al. we considered overlaps between bins and regions defined by functional genomic elements when creating the predictor space of our models. If a elemental region was located completely within a bin then the predictor ($X_{i}$) was labeled as 1. If a bin did not overlap with any region then it was labeled as 0. In the case of partial overlaps, a fraction denoting the percent overlap from region to bin was used. In the event of multiple regions within the same bin. The sum total of fractions or instances of regions was used as in Figure X.

@@

We developed a pipeline that systematically evaluates the problems associated with TAD boundary data in a predictive modeling setting (Figure X). The pipeline addresses X key questions in aiming to improve predictive performances and accurately identify which functional genomic elements are most predictive of TAD boundaries. Prior to initiating the pipeline, predictors with near-zero variances were removed. To identify said predictors, the nearZeroVar function in the CARET package was used. The function considers two metrics. First, the ratio of the most frequent value over the second most frequent value was calculated. Ideally, this ratio would be close to one for well-behaved predictors. Next, the number of unique values divided by the total number of samples multiplied by 100 was calculated. Here, if the frequency ratio was greater than 20 and the percentage of unique values was less than 10, the predictor was flagged as having near-zero variance and removed.

### Question 1: CLASS IMBALANCE

How severely unbalanced were the classes and what was the affect on model performance? When binning the genome, the sparcity of TAD boundaries contributed to heavily imbalanced classes. That is, proportionally, only a small number of genomic bins overlapped with a flanked TAD boundary. To assess this issue two multiple logistic regression models were performed. One with unbalanced classes and one with balanced classes created by random under-sampling from the majority class. It was found that the model with balanced classes out-performed the model with unbalanced classes. However, given the vast disparity between classes simply performing on iteration of random under-sampling is not sufficient in determining the best possible balancing technique.

### Question 2: BALANCING TECHNIQUE

Which balancing technique performs best? To assess this issue, three methods were used. The first consisted of performing X iterations of over-sampling from the minority class to match the majority class. At each iteration, a multiple logistic regression model was used and performance metrics were recorded. At the end of the X iterations, the performance metrics were aggregated by taking the average. The second method consisted of random under-sampling. Similar to over-sampling, the majority class was randomly under-sampled over X iterations to create balanced classes. Again, at each step a multiple logistic regression model was performed and the results were aggregated using the average.  The last method incorporated the SMOTE command from the DmWR package in R. SMOTE stands for Synthetic Minority Over-sampling Technique. The SMOTE algorithm incorporates both under-sampling (percent under) of the majority class and over-sampling (percent over) of the minority class. The minority class is over-sampled by taking each minority class sample and introducing synthetic examples along the line segments joining any/all of the k minority class nearest neighbors (here, k=5). Then, the majority class is under sampled by taking a random sample that matches some pre-specified percentage of the updated minority class. A total of four different combinations of percent over and under-sampling were used which included: 100/200, 200/200, 300/200, and 400/200. It is important to note that using a combination of 100/200 (percent over/percent under-sample) will always result in perfectly balanced classes. The other combinations created unbalanced classes with the negative class as the majority to reflect the proportionality of the classes from the original data. Multiple logistic regression models were used and performance metrics were assessed and compared with each of the other techniques. It was determined that the model with data untilizing random under-sampling out-performed all other other balancing techniques. However, to this point we only considered overlaps between genomic bins and the regions that defined functional genomic elements.

### Question 3: DISTANCE PREDICTORS

What affect does the inclusion of distances from bin to region have on model performance? To this point we had only considered overlaps between genomic bins and the regions that defined functional genomic elements. Now we added another type of predictor created from each genomic element. This consisted of the distance from the center of each genomic bin to the center of the nearest region defining a genomic element as depicted in Figure X. To assess any improvement with the inclusion of these predictors we compared the performances of three multiple logistic regression models: one with only the overlap predictors, one with only the aforementioned distance predictors, and one with both.  It was found that the inclusion of distance predictors improved model performance.  Likewise, the model with both types of predictors futher improved model performance. As is common, with genomic data, the use of continuous predictors (especially with regards to distances) results in highly right skewed data with distances ranging from 0 to tens of thousands in base pairs. This can affect the convergence of likelihood maximization.

### Question 4: NORMALIZATION

What normalization technique best improves model performance?
